
\documentclass{article}

\usepackage{amsmath}
\usepackage{latexsym,amsthm,amssymb,amscd,url,enumerate}
\usepackage[show]{ed}
\usepackage[all]{xy}
\usepackage{tikz}
\usepackage{tikz-cd}
\usepackage{leftidx}
\usetikzlibrary{arrows}

\begin{document}

\newcommand{\E}{\mathtt{e}}
\newcommand{\B}{\mathtt{b}}
\newcommand{\C}{\mathtt{c}}
\newcommand{\R}{\mathtt{R}}
\newcommand{\LL}{\mathtt{L}}
\newcommand{\Sem}{\mathtt{S}}
\newcommand{\true}{\mathtt{true}}
\newcommand{\false}{\mathtt{false}}
\newcommand{\andsym}{\mathtt{and}}
\newcommand{\orsym}{\mathtt{or}}
\newcommand{\notsym}{\mathop{\mathtt{not}}}
\newcommand{\ifsym}{\mathtt{if}}
\newcommand{\then}{\mathtt{then}}
\newcommand{\elsesym}{\mathtt{else}}
\newcommand{\whilesym}{\mathtt{while}}
\newcommand{\dosym}{\mathtt{do}}
\newcommand{\skipsym}{\mathtt{skip}}
\newcommand{\nil}{\mathtt{nil}}
\newcommand{\case}{\mathtt{case}}
\newcommand{\digit}{\mathtt{d}}
\newcommand{\negation}{\mathtt{neg}}
\newcommand{\Digit}{\mathbf{Digit}}
\newcommand{\denot}[1]{\mathtt{[[{#1}]]}}

\newtheorem{theorem}{Theorem}						
\newtheorem{definition}[theorem]{Definition}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{axiom}{Axiom}
\newtheorem{notation}[theorem]{Notation}
\newtheorem{corollary}[theorem]{Corollary}

\title{\textbf{15-812 Semantics of Programming Languages \\ Spring 2015 \\ {\large Assignment 2}}}
\date{}
\author{}

\maketitle

\noindent In this assignment we will revisit the imperative language $\mathcal{L}$ defined by the syntax
\begin{align*}
& \E := \mathtt{n} \; | \; \mathtt{x} \; | \; \E_1 + \E_2 \; | \; \E_1 \cdot \E_2 \; | \; \negation(\E) \; | \; \ifsym \; \B \; \then \; \E_1 \; \elsesym \; \E_2 \\ 
& \B := \true \; | \; \false \; | \; \B_1 \; \andsym \; \B_2 \; | \; \B_1 \; \orsym \; \B_2 \; | \; \notsym \; \B \; | \; \E_1 \leq \E_2 \\
& \C := \skipsym \; | \; \mathtt{x} := \E \; | \; \C_1 ; \C_2 \; | \; \ifsym \; \B \; \then \; \C_1 \; \elsesym \; \C_2 \; | \; \whilesym \; \B \; \dosym \; \C
\end{align*}
and the state-transition $[[\cdot]]$, input/output $\mathcal{I}$, and termination $\mathcal{T}$ semantics for $\mathcal{L}$ as defined in class. Recall that two programs $\C_1, \C_2$ are 
\begin{enumerate}
\item \emph{equivalent} with respect to a semantics $\Sem$ if $\Sem(\C_1) = \Sem(\C_2)$
\item \emph{contextually equivalent} with respect to $\Sem$ if $\forall \mathcal{C}, \Sem(\mathcal{C}[\C_1]) = \Sem(\mathcal{C}[\C_2])$, where $\mathcal{C}$ ranges over program contexts
\end{enumerate}
For two semantics $\Sem_1, \Sem_2$ we say that
\begin{enumerate}
	\item $\Sem_1$ is \emph{compositional} with respect to $\Sem_2$ if for any two programs $\C_1,\C_2$, we have that $\Sem_1(\C_1) = \Sem_1(\C_2)$ implies $\forall \mathcal{C}, \Sem_2(\mathcal{C}[\C_1]) = \Sem_2(\mathcal{C}[\C_2])$
	\item $\Sem_1$ is \emph{fully abstract} with respect to $\Sem_2$ if for any two programs $\C_1,\C_2$, we have that $\forall \mathcal{C}, \Sem_2(\mathcal{C}[\C_1]) = \Sem_2(\mathcal{C}[\C_2])$ implies $\Sem_1(\C_1) = \Sem_1(\C_2)$
	\end{enumerate}
	A semantics $\Sem$ is \emph{compositional} if it is compositional with respect to itself.\bigskip

\noindent \textbf{Please justify all answers.}

\section{Revisiting semantics}
\begin{enumerate}
\item[1.1] 
\begin{enumerate}
	\item[a)] Is the input/output semantics $\mathcal{I}$ compositional? Is it fully abstract with respect to the termination semantics $\mathcal{T}$?
  \item[b)] Is the termination semantics $\mathcal{T}$ compositional? Is it fully abstract with respect to the input/output semantics $\mathcal{I}$?
\end{enumerate}
\item[1.2]
\begin{enumerate}
	\item[a)] Give a semantics $\Sem$ such that the state transition semantics $[[\cdot]]$ is not compositional with respect to $\Sem$.
	\item[b)] Give a semantics $\Sem$ such that the state transition semantics $[[\cdot]]$ is not fully abstract with respect to $\Sem$.
	\end{enumerate}

\item[1.3] In the previous homework assignment, question 2, you defined a denotational semantics $\Sem$ for the extended imperative language $\mathcal{L}_e$:
\begin{align*}
& \E := \mathtt{n} \; | \; \mathtt{x} \; | \; \E_1 + \E_2 \; | \; \E_1 \cdot \E_2 \; | \; \negation(\E) \; | \; \ifsym \; \B \; \then \; \E_1 \; \elsesym \; \E_2 \\ 
& \B := \true \; | \; \false \; | \; \B_1 \; \andsym \; \B_2 \; | \; \B_1 \; \orsym \; \B_2 \; | \; \notsym \; \B \; | \; \E_1 \leq \E_2 \\
& {\color{red} \LL := \nil \; | \; (\mathtt{n},\C)::\LL} \\
& \C := \skipsym \; | \; \mathtt{x} := \E \; | \; \C_1 ; \C_2 \; | \; \ifsym \; \B \; \then \; \C_1 \; \elsesym \; \C_2 \; | \; \whilesym \; \B \; \dosym \; \C \; | \; {\color{red} \case \; \E \; \LL \; \C}
\end{align*}
Is $\Sem$ compositional? Is the state transition semantics $[[\cdot]]$ compositional with respect to $\Sem$? Is it fully abstract?
\end{enumerate}

\section{Imperative reasoning}
\begin{enumerate}

\item[2.1] Prove or disprove the following statements.
\begin{enumerate}
\item[a)] The programs $(\whilesym\; \B \; \dosym \; \C); (\ifsym \; \B \; \then \; \C_1 \; \elsesym \; \C_2)$ and $(\whilesym \; \B \; \dosym \; \C);\C_2$
are equivalent with respect to the state transition semantics $[[\cdot]]$.
\item[b)] The programs $\whilesym\; \B \; \dosym \; \C$ and $\whilesym \; \B \; \dosym \; (\C;\ifsym \; \B \; \then \; \C \; \elsesym \; \skipsym)$ are equivalent with respect to the state transition semantics $[[\cdot]]$.
\item[c)] The programs $\whilesym\; \B \; \dosym \; (\whilesym \; \B \; \dosym \; \C)$ and $\whilesym\; \B \; \dosym \; \C$ are contextually equivalent with respect to the state transition semantics $[[\cdot]]$.
\end{enumerate}

\item[2.2] Under what conditions are the programs
\begin{enumerate} 
\item[a)] $\C;(\ifsym \; \B \; \then\; \C_1\; \elsesym \; \C_2)$ and $\ifsym \; \B \; \then\; (\C; \C_1) \; \elsesym \; (\C; \C_2)$
\item[b)] $(\ifsym \; \B \; \then\; \C_1 \; \elsesym \; \C_2); (\ifsym \; \B \; \then \; \C_3 \; \elsesym \; \C_4)$ and $\ifsym \; \B \; \then\; (\C_1;\C_3) \; \elsesym \; (\C_2;\C_4)$
\end{enumerate}
equivalent with respect to the state transition semantics $[[\cdot]]$?

\item[2.3] Show that the following program computes the greatest common divisor of two positive integers $x,y$:
\begin{align*}
& \whilesym \; \notsym(x = y) \; \dosym \\
& \;\;\;\;\; \whilesym \; (x < y) \; \dosym \; y := y - x; \\
& \;\;\;\;\; \whilesym \; (y < x) \; \dosym \; x := x - y
\end{align*}
Here we write $\E_1 < \E_2$ for $\E_1 \leq \E_2 \; \andsym \; \notsym(\E_1 = \E_2)$ and $\E_1 - \E_2$ for $\E_1 + \negation(\E_2)$. What are the values of the above program under the input/output and termination semantics? How do your answers change if we instead consider the program
\begin{align*}
& \whilesym \; \notsym(x = y) \; \dosym \\
& \;\;\;\;\; \whilesym \; {\color{red} (x \leq y)} \; \dosym \; y := y - x; \\
& \;\;\;\;\; \whilesym \; {\color{red} (y \leq x)} \; \dosym \; x := x - y
\end{align*}
\end{enumerate}
\end{document}


